---
title: "Mathematical notation for plots"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 1, fig.width = 5)
library("learnr")
library("tidyverse")
theme_set(theme_classic())

tutorial_options(exercise.cap = "Exercise")
g <- ggplot()
```

## Introduction

When making a publication ready plot, we often need to use special characters and mathematical notation  for the axes and other labels.

 - $TP~concentration~\mu g~l^{-1}$
 - $Conductivity~\mu S~cm^{-1}$
 - $\sqrt{Standardised~residuals}$
 - $\delta^{15}N$
 
The functions `expression()` and `bquote()` are powerful tools for annotating figures with mathematical notation in R. 
This functionality is not obvious from their respective help files. 
`demo(plotmath)` and `?plotmath` shows the huge potential of `expression()`, but do not have many examples relevant to biologists with graphs to label.

Expressions are comprised of symbols. 
The expression $TP ~ concentration ~\mu g~l^{-1}$ includes the symbols $TP$, $concentration$, $\mu$, $g$, $l$, and $^{-1}$. 
In this tutorial, you will learn how to generate these symbols and combine them with `expression()`.

The key to getting expressions to work is to start with something simple, test it and edit until it works, then add the next symbol and repeat. 
This is the approach we are going to use here.

## `expression()`

I’m going to use `expression()` with `labs()`, but the same expressions can be used with some other functions that write text on a plot (including base plots). 
(but see section on parsing below)

The simplest use of expression is take a symbol that is a single word or number.

```{r}
g <- ggplot() + #empty plot
  labs(x = expression(TP))
g
```

Notice that `expression()` does not require quote marks. 
This first example of  `expression()` is entirely pointless except as a starting point for something more complex.

If we want two symbols with a space between them, we need to separate them with a tilde `~`. 
If we don't want a space, we can separate symbols with `*` (if you actually want a `*`, use `%*%`).

```{r}
g + labs(x = expression(TP~concentration))
```

Some symbols have special meanings, for example, to get a Greek letter, write its name in lower case. 
Capitalise the first letter to get an upper-case Greek letter. 
If for some reason you want to have "beta" written on your plot, it must be in quotes.

```{r}
g + labs(x = expression(TP~concentration~mu*g))
```

Subscript or superscript can be added to a string using `^` and `[]` notation respectively. 
If you want more than one symbol to be in superscript, you need to enclose them in braces `{}`.

```{r}
g + labs(x = expression(TP~concentration~mu*g~l^-1))
```

To start an `expression()` with a superscript (or subscript), put `""` first.

```{r, eval = FALSE}
g + labs(x = expression(""^14*C~years~BP))
```



```{r equations, echo = FALSE, message=FALSE}
library(kableExtra)
read_csv('
What, How, Output         
"Lower-case Greek letters", "delta", "$\\delta$"
"Upper-Case Greek Letters", "Delta", "$\\Delta$"
"Subscript", "CO[2]", "$CO_{2}$"
"Superscript", "m^{-2}", "$m^{-2}$"
"Italics", "italic(Navicula)", "_Navicula_"'
) %>% 
  kbl(booktabs = TRUE, caption = "Some common notation") %>% 
  kable_styling(full_width = TRUE)
```


### Your turn

Make the x-axis label read $Area~m^2$.

```{r expression1, exercise=TRUE}
ggplot() + 
  labs(x = ___)
```

```{r expression1-solution}
ggplot() +
  labs(x = expression(Area~m^2))
```

Make the x-axis label read $Flux~g~m^{-2}~yr^{-1}$.

```{r expression2, exercise=TRUE}
ggplot() + 
  labs(x = ___)
```

```{r expression2-solution}
ggplot() +
  labs(x = expression(Flux~g~m^-2~yr^-1))
```

Make the x-axis label read $\delta^{15}N$.

```{r expression3, exercise=TRUE}
ggplot() + 
  labs(x = ___)
```

```{r expression3-solution}
ggplot() +
  labs(x = expression(delta^15*N))
```

Make the x-axis label read _Aulacoseira granulata_ var. _angustissima_.

```{r expression4, exercise=TRUE}
ggplot() + 
  labs(x = ___)
```

```{r expression4-solution}
ggplot() +
  labs(x = expression(italic(Aulacoseira~granulata)~var.~italic(angustissima)))
```

## Parsing

Some functions such as `geom_text()`, `facet_wrap` and `scale_x_discrete` do not accept expressions directly. 
Instead, they take can take parse a string and convert it to an expression.
Here is a minimal example that uses parse in three different ways to add sub scripts and superscripts to text, axis labels and facet strips. 

```{r parsed, fig.height = 3, fig.width=4}
dat <- tibble(x = c("NO[3]^-1", "SO[4]^-2"),
                  y = c(10, 8)) |> 
  mutate(x = factor(x))

ggplot(dat, aes(x = x, y = y)) +
  geom_col() +
  scale_x_discrete(labels = parse(text = levels(dat$x))) +
  geom_text(aes(label = x), 
            parse = TRUE, 
            nudge_y = 1) + 
  facet_wrap(~ x, labeller = "label_parsed")
```




## Including objects

So far so good. But sometimes, we want to include the value of an R-object in the expression.

For example, if we wanted to label a point with its `x` value, this will not work.

```{r, error = TRUE}
value <- 5
g + labs(x = expression(expression(x==value))
```

Fortunately, there is another function, `bquote()`, which will let us do this if we use the helper function `.()` with the object we want to include.


```{r}
x <- 5
g + labs(x = bquote(x==.(x)))
g + labs(x = bquote(x==.(x)~mu*g~l^-1))

```


### Other tips

- `demo(plotmath)` and `?plotmath` are very useful once you have understood the basics.
- Some symbols are not defined by `plotmath` can be included using [unicode](http://www.unicode.org/charts/). 
For example a per mil symbol (‰) - `\u2030` 
- [LaTeX equations](https://biostats-r.github.io/biostats/rmarkdown/figures-tables-and-equations.html#equations) can be converted to expressions with `latex2exp::TeX`.



If you understand these examples, you should be able to use the remainder of the functionality demonstrated by `demo(plotmath)` and at `?plotmath`.