---
title: "Working with dates and times"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, rows.print = 3)
library("learnr")
library("tidyverse")
library("lubridate")
library("here")
library("conflicted")
conflict_prefer(name = "here", winner = "here")
conflict_prefer(name = "filter", winner = "dplyr")

tutorial_options(exercise.cap = "Exercise")

temperature <- read_csv(here("data", "Florida_2020-04-20_2020-04-23_1589018968.csv"))
```

## Date and time formats

Converting strings to dates is tricky.

Many formats
 
- 5th December 2020
- 12-5-2020
- 5.12.20
- 5 Dec 20
- 2020/12/5

Possibly languages: January, januar, janvier, 一月

When entering data into a spreadsheet or similar, I strongly recommend using the [international standard format](https://en.wikipedia.org/wiki/ISO_8601) for dates and times.

- Year-Month-Day Hour:Minute:Second
- 2020-12-5 12:23:13

## Converting dates and times

It is usually more convenient to have dates in a `Date` format and date/times in a `POSIXct` format rather than as a character.
This allows dates to be manipulated, sorted, and plotted properly.

This can be done using base R.

```{r base-R-convert}
dat <- "26 October 2016 14:39:10 CEST" #date with time-zone
as.POSIXct(dat, format = "%d %B %Y %H:%M:%S")
```

Here the `%d` represents the day, `%B` represents the month written as the full word (`%b` would work if the month was the abbreviated) and `%Y` represents the 4-digit year. 
There is a complete alphabet of codes you can use. 
See `?strptime` for the complete list.

## Converting dates and times with `lubridate`

The `lubridate` package makes converting dates much easier.

### No more `"%d %B %y"`

With `lubridate`, you don't need to remember the alphabet of codes and get the format **exactly** right, you just need to select the function with the day, month and year in the correct order.

```{r lubridate-convert}
library("lubridate")
dat <- "26 October 2016 14:39:10"
dmy_hms(dat)
```

`dmy_hms` expects the order to be days-months-years followed by hours:minutes:seconds. 
Provided the elements are in this order, any format can be used.

```{r lubridate-dmy_hms}
dmy_hms("26-10-16 14.39.10")
dmy_hms("26th Oct 2016 14 39 10")
dmy_hms("261016143910")
```


### Different formats

There are a load of other functions in the `lubridate` package for coping with dates with elements in different orders.

For example

```{r lubridate-formats}
dmy_hm("26th Oct 2016 14.39") # no seconds
ydm_hms("2016 26th Oct 14 39 10") # year first then day
mdy_hms("Oct 26th 2016 14 39 10") # month first (US standard)
```

Look at the help file for more function and arguments.

If you have an invalid format or an impossible date, you will get a warning.

```{r bad-date}
dmy("28-F-2020") # Invalid format - use February, Feb or 2
dmy("31-Feb-2020") # Impossible date
```


### Your turn

Use the `lubridate` package to convert the following to date or datetime format.


```{r lubridate, exercise = TRUE}
date1 <- "28/02/1999"

date2 <- "July 1 2001 2:14"
```

```{r lubridate-solution}
date1 <- "28/02/1999"
dmy(date1)

date2 <- "July 1 2001 2:14"
mdy_hm(date2)
```

## Date arithmetic

Dates and times can be used in calculations

```{r date-arithmetic}
ymd("2020-5-18") - 5 #base unit is days for dates
ymd("2020-5-18") - period(5, unit = "months")
ymd("2020-5-18") - ymd("2020-5-13")
```


### Your turn

Find the number of days until Christmas.

```{r christmas, exercise = TRUE}
#How many days until Christmas 2020

```

```{r christmas-solution}
ymd("2020-12-25") - today()
```

## Extracting elements of the date

Given a date or time, we can extract different elements, for example the month, day or hour.

```{r}
month(today())
yday(today())# day of year
```

## Languages

R uses locales to know what language to expect the date to the written in. See `?locales`

You can change the locale to Bokmål for the whole session with 

```{r date-locales}
Sys.setlocale("LC_TIME", "nb_NO.utf8")
dmy("1 januar 2020")
```

Or use the `locale` argument of `dmy`.

To see the available locales on your computer, run 

```{r available-locales, eval = FALSE}
system("locale -a", intern = TRUE)
```

```{r date-locale-revert, echo = FALSE, include = FALSE}
Sys.setlocale("LC_TIME", "en_DK.utf8")
```



## Dates and `dyplr`

As an ecologist, I spend a lot of time processing climate data  for comparison with ecological data.
The data are typically in R in a data.frame, so `dplyr` is useful for manipulating them.

Dataset `temperature` includes a few days of air temperature from the [GFI weather station](https://veret.gfi.uib.no/), Bergen.

The dataset has three columns, `Dato`, `Tid` and `Lufttemperatur`. I have imported the data with `readr::read_csv` so the first two columns have automatically been converted to date and time formats.

```{r print-temperature}
temperature
```

### Using `mutate` to make a new column

I want to make a single date-time column. To do this I need to combine the data and time columns with `paste`, and then use a `lubridate` function for the conversion. 

```{r date-time-temperature}
temperature <- temperature %>% 
  mutate(
    date_time = paste(Dato, Tid), #combine
    date_time = ymd_hms(date_time)#convert
    ) 

temperature
```

### Using `filter` to select rows

I often need to `filter` data to remove bad data or restrict the dates and times in the dataset.

For example, calculate the mean temperature of the afternoon of the 21st April, we first need to `filter` the data. 
`between` is a useful helper function here to avoid having to write a more complicated logical condition.

```{r filter-temperature}
temperature %>% 
  filter(between(date_time, 
                 left = ymd_hm("2020-04-21 12:00"),
                 right = ymd_hm("2020-04-21 17:00"))) %>% 
  summarise(mean_temp = mean(Lufttemperatur))

```



### Using `group_by` and `sumararise`

I often want to summarise climate data to calculate mean monthly temperatures. We can do an analogous task with the `temperature` dataset to calculate mean hourly temperatures.

```{r mean-hourly}
temperature %>% 
  mutate(hour = hour(date_time)) %>% 
  group_by(hour) %>% 
  summarise(mean_temperature = mean(Lufttemperatur))
```



```{r filter, exercise=TRUE}
# Change the filter to select February rather than January
```

## Plotting dates

We can plot the temperature data with `ggplot`.

```{r}
ggplot(temperature, aes(x = date_time, y = Lufttemperatur)) + 
  geom_line()
```

`ggplot` recognises that the x-axis variable is a date-time and plots the axis with appropriate labels. 
The language will depend on the locale (see `Sys.getlocale()`)


```{r}
ggplot(temperature, aes(x = Tid, y = Lufttemperatur, group = Dato)) + 
  geom_line()
```

```{r}
ggplot(temperature, aes(x = Tid, y = Lufttemperatur, group = Dato)) + 
  geom_line() + 
  scale_x_time(breaks = seq(0, 24, 6) * 3600, labels = scales::time_format("%H:%M"))
```