---
title: "Importing data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Importing data

R can import data in many formats, including

 - text files with delimiters between columns (e.g csv files)
 - excel files
 
Which this tutorial covers, but also databases (with the `DBI` package), and other formats with the `haven` package.

Best practice is to leave the data in the format you get it rather than trying to convert it to some other format.

## File paths

R needs to be able to find your data files.

Usually the best way to do this is to use paths relative to your RStudio projects working directory.

The problem with using an absolute path, for example,

```
c:/my_directory/my_project/data/my_file.csv
```

Is that this makes it difficult to collaborate as collaborators may want to have the data in a different place.

Much better is to put the data into a folder called `data` in your Rstudio project, and use a relative path.

```
data/my_file.csv
```

If you share your entire project, it will work without all the paths needing to be changed.

### `here`

Best practice is to use the `here` package to keep track of your working directory. 
`here` will let your code work for people who do not use RStudio.

```{r here, eval = FALSE}
library(here)
here("data", "my_file.csv")
```

### `setwd()`

You probably never need to use `setwd()` if you have your RStudio project set up correctly.

## Data organisation in spreadsheets

Some data are easier to import than others.
These are some recommendations from Broman and Woo (2018) that will make data easy to import.

- be consistent, write dates like YYYY-MM-DD
- do not leave any cells empty
- put just one thing in a cell
- organize the data as a single rectangle (with subjects as rows and variables as columns, and with a single header row) 
- create a data dictionary
- do not include calculations in the raw data files
- do not use font colour or highlighting as data
- choose good names for things

Broman, K. W. & Woo, K. H. (2018) Data Organization in Spreadsheets, The American Statistician, [72:1, 2-10,](https://doi.org/10.1080/00031305.2017.1375989)

## Importing text files with `readr`
 
Text files with delimiters are probably the easiest type of data to import.

Common delimiters are

 - commas (csv)
 - semi-colons (csv)
 - tabs (txt or tab)
 
 Open the file in a a text editor (NOT excel), to see what type of delimiter you have, and also check what the decimal separator is.
 
Now you can import the file.


```{r readr, eval = FALSE}
library(readr) # part of tidyverse. not needed if tidyverse already loaded.

my_data <- read_delim(file = "data/my_file.csv", delim = ",")
```

TIP: once you have typed the quotation marks, press Tab and RStudio will suggest possible files.

`read_delim` has many arguments. 
In addition to `file` and `delim`, some of the most useful are

- `locale` : use this to change the decimal separator with `locale = locale(decimal_mark = ",")`
- `skip` : skips some lines at the top of the file. 
Useful if the file starts with some metadata.
- `col_types` : `read_delim` will guess what type of data is in each column and convert it accordingly. 
Sometimes it will make a mistake. 
`col_types` will force it to use the correct type.

See the help file for other arguments.

The advantages of `read_delim` over the base R `read.table` and related functions are that it is faster, and gives information on problems in the data.

## Importing Excel files with `readxl` 
 
Excel (and other spreadsheet software) is convenient for entering data but xlsx and xls files are not the ideal format to store data in.
However, if you are given/download data in Excel format, you should leave it in that format and import it directly to R.

There are several packages that let you import excel files. 
I usually find `readxl` to be the best (some of the others need Java installing - not always a fun experience - or are very slow)

```{r read_excel, eval = FALSE}
library(readxl) # installed with tidyverse, but NOT loaded with library(tidyverse)
file <- "myfile.xls"
# find the names of the excel sheets
excel_sheets(path = file)

#import the desired sheet - imports the first by default
my_data <- read_excel(path = file, sheet = "Sheet1")
```
 
As with `read_delim`, `read_excel` has arguments to `skip` rows at the top of the file, or force the `col_types`.

## Importing multiples files

Sometimes you have multiple files to import, that have the same format. 
For example, 12 files each containing weather data for one month.
Of course, you can import them one-by-one, but that gets tedious and error prone once there are more than a few files.

Instead we can use the `purrr` package to iterate over multiple files (or equivalently sheets in an excel file).

Here is a minimal example

```{r, eval = FALSE}
library(tidyverse)#purrr is part of tidyverse
list.files(path = "data", pattern = "\\.csv$", full.names = TRUE) %>% 
  map_dfr(~read_delim(file = .x, delim = ","))
```

`list.files` will list all the csv files in the data directory in a vector.
The pipe passes this to `map_dfr` which will iterate over the vector, importing files one at a time and combining all the files into one large data frame.
 
## Creating small data sets with code

Sometimes you need a small dataset (few rows and columns) that you don't already have in a file.
You can create these directly in R.

### With `readr::read_delim`

Instead of giving `read_delim` the file name of a data file, you can give it a block of text to read directly.


```{r readr-textblock}
svalbard_islands <-  read_delim("island, Latitude, Longitude
Nordaustlandet, 79.558405, 24.017351
Prins Karls Forland, 78.554090, 11.256545", delim = ",")
```

Don't leave any extra space in the text block or it can be included in the output.

### With `tibble::tribble`

Or you can use `tribble()`.
`tribble` takes the column names preceded by a tilde, and then the data, all separated by commas.
Anything other than a number need to be enclosed in quotation marks.


```{r}
svalbard_islands <-  tribble( ~ island, ~ Latitude, ~ Longitude,
   "Nordaustlandet", 79.558405, 24.017351,
   "Prins Karls Forland", 78.554090, 11.256545)
```

You don't need to format it so that each row has it's own line but it makes it easier to read.

It is largely a matter of personal preference whether you want to use `read_delim` or `tribble`.